### 2012-10-22 3.tex
edits on 3

There are several arguments against this inference. First, linguists have long noted statistical generalizations which have a diachronic cause but which lack any synchronic reality. One of the earliest arguments of this type was put forth by Saussure (\citeyear[202f.]{CLG}), who recognizes the a diachronic process which changed Old Latin intervocalic \emph{s} to \emph{r}, and that Classical Latin intervocalic \emph{s} is rare as a result. However, Saussure denies that this generalization has any status in Classical Latin.\footnote{See \citet{Gorman2012e} and citations therein for a full explication of Saussure's position on rhotacism, and a restatement of the phonological argument in generative terms.} Similarly, [ʃ] in Modern English derives from Old English [sk] (e.g., \emph{fisc} 'fish'), and since long vowels before complex codas are not found in Old English, [ʃ] is rarely preceded by long vowels in word-final syllables to this day. \citet{Iverson2005} label this constraint ``phonologically accidental'', since a millennium of speakers have produced coinages (e.g., \emph{posh}, \emph{swoosh}) and borrowings (e.g., \emph{douche}, \emph{capiche}) in utter disregard of this generalization. Secondly, wordlikeness generalizations are absolutely constrained by negative results in learnability (see \citealt{Yang2012} for a recent review). Finally, there are a number of results which suggest patterns which were obvious to the linguist go unlearned \citep[e.g.,][]{Becker2011,Hayes2006,Hayes2009,HayesInPress}. This argues against a direct link between statistical criteria and the synchronic reality of phonotactic generalizations. The danger is that grammatical models which closely mimic lexical statistics but which lack psycholinguistic support (e.g., Gitksan: \citealt{Brown2010}; Muna: \citealt{Anttila2008a}, \citealt{Coetzee2008a}; Shona: \citealt[][385]{Hayes2008a}) may not match speakers' knowledge.

This chapter is a comprehensive investigation of three well-known phonotactic generalizations in Turkish using lexical statistics and the results of a wordlikeness task performed by \citet{Zimmer1969}. \citeauthor{Zimmer1969}'s experiments merit reconsideration because there is widespread disagreement concerning the results and prior discussions fail to relate the behavioral data to lexical statistics and to competing formalizations of the generalizations involved.

@@

Two less familiar notations are used in this chapter. First, directional application conditions are used to derive the left-to-right, iterative properties of. Directional application explains provides a panacea for the considerable body of evidence against simultaneous rule application adduced in the 1970s (e.g., \citealt[209f.]{Anderson1974}, \citealt{Howard1972}, \citeauthor{Kenstowicz1973} \citeyear[14f.]{Kenstowicz1973}, \citeyear[189f.]{Kenstowicz1977}, \citeyear[318f.]{Kenstowicz1979}, \citealt{Piggott1975}). In fact, harmony in Turkish is one such case, insofar as simultaneous application makes the wrong predictions. Further, \citet{Johnson1972} and \citet{Kaplan1994} prove that rules specified for directional application are no more complex in the terms of formal learnability than that do not. Secondly, rather than the use of an unbounded number of Greek-letter variables ($\alpha$, $\beta$, etc.) over feature values $\{+, -\}$, only a single variable, denoted by `$=$', is used. Following \citet{McCawley1973}, a structural description [$=$F]\ldots{}[$=$F] matches a string S$_i$\ldots{}S$_j$ if and only if S$_i$ and S$_j$ are both [$+$F] or both [$-$F].\footnote{This is more restrictive than the Greek-letter variable notation, in that the value of $=$ for some feature F cannot be applied to F$'$ if F $\ne$ F$'$; \citet{Odden2012} dismisses well-known counterexamples against this restriction adduced in \emph{SPE} (352-353).}

@@

It is necessary to dispense with an alternative proposed by \citet{Clements1982} and \citet{Inkelas1997} which reintroduces this duplication. Root vowels exhibit a robust contrast for backness (e.g., \emph{kül} `ash' vs.  \emph{kul} `servant', \emph{kepek} `bran' vs. \emph{kapak} `lid'), whereas harmonic roots are those in which the backness of any remaining vowels is predictable. \citeauthor{Clements1982} propose that these vowels, as well as harmonizing suffix vowels, are underspecified for backness, whereas the non-initial vowels of disharmonic roots are fully specified. This is schematized below.

@@

While harmony in non-derived environments can be inferred from the aforementioned suffix alternations, no evidence has yet been presented to show internalize the tendency for roots to conform to backness harmony. If Turkish speakers do not attend to this generalization, there is no need for the grammar to account for it. Several other ``external'' facts suggest  that this is not the case, with the caveat that uncritical acceptance of hypotheses linking loanword adaptation and language games to the grammar is no way being implied.

@@

A number of studies have investigated the role of harmony in word-spotting tasks, though to mimic auditory word recognition and segmentation in natural settings. Many of these studies have been carried out in Finnish, which has a vowel harmony system similar to that of Turkish. \citet{Suomi1997} and \citet{Vroomen1998} task Finnish speakers with identifying harmonic disyllables in an auditory stream. When the syllable preceding the disyllabic target has a different backness specification than the target, this faciliates speakers' recognition abilities. Presumably, disharmony facilitates the recognition of word boundaries. \citet{Kabak2010} find that Turkish \textsc{Backness Harmony} has a a similar effect: Turkish speakers are quicker and more accurate at the task of spotting the nonce target word \emph{pavo} when preceded by a disharmonic juncture (e.g., \emph{gölü-PAVO}) than when preceded by a harmonic juncture (e.g., \emph{golu-PAVO}). \citet{Kabak2010} find that this effect does not obtain for speakers of French, a language which lacks vowel harmony. This implies that Turkish speakers have internalized the tendency of harmonic sequences to be root-internal and of disharmonic transitions to cross word boundaries.

@@

\noindent 
This claim yields itself naturally to the Fisher exact test. Let P denote a sequence of one or more consonants, one of which is labial, and let T denote a sequence of one or more consonants none of which is labial. The null hypothesis is that \emph{a}P\emph{u} sequences, with P representing a cluster containing a labial consonant, which conform to \textsc{Labial Attraction}, are no more likely than would be expected from other \emph{a}T\emph{u} sequences violating \textsc{Roundness Harmony}. The resulting counts are shown in Table \ref{las}. Whereas the sequence \emph{a}P\emph{u} is more than twice as likely as \emph{a}P\emph{ı}, the sequence \emph{a}T\emph{u} is four times less likely than \emph{a}T\emph{ı}. This is a significant interaction predicted by \textsc{Labial Harmony}, but contradicting a statistical interpretation of the claim by \citeauthor{Clements1982}.

@@

Both of the experiments performed by \citet{Zimmer1969} include 5 pairs which differ on in whether or not the nonce words conform to, or violate, \textsc{Backness Harmony}. As can be seen from Table \ref{bhw}, harmonic pairs are preferred approximately 6-to-1, and aggregating over speakers, no disharmonic member of a pair is favored. Speakers have a highly reliable preference for nonce words which exhibit \textsc{Backness Harmony} ($\gamma = 0.694$, $p = 1.7$\e{-59}). It is interesting to note that the disharmonic nonce word which has the highest rating is found in the pair \emph{terüz}-\emph{teruz}, both of which violate \textsc{Roundness Harmony}. While this is little more than an anecdote, this may be  indicative of a link between the two processes, and their exceptions, in the minds of native speakers.

@@

\citet{NiChiosain1993} and \citet{Ito1995b} argue that \textsc{Labial Attraction} is true only of native vocabulary. This would be a potential confound for the experimental results, since it is not implausible that the subjects' in \citeauthor{Zimmer1969}'s study treated nonce words as if they were loanwords. 
%\footnote{Indeed, many wordlikeness studies include instructions to the participants to treat the stimuli much as if they were loanwords \citep[e.g.,][]{Hay2004a}. This may influence the choice of task model used by participants.}
This was not the opinion of \citet[266]{Lees1966a}, and it is conclusively refuted by  \citet{Inkelas2001}. They observe that in the subset of the TELL database with etymological coding, a full 75\% of foreign vocabulary conform to \textsc{Labial Attraction}, which is surprisingly a smaller percentage than that of the foreign vocabulary (52\%), a significant trend (Fisher exact test, $p = 0.042$). One plausible explanation is that many of the languages in contact with Turkish lack the \emph{ı} phoneme needed to violate \textsc{Labial Attraction}.

@@

It has been shown that statistical reliability is not a sufficient condition for a sequence structure generalization to be reflected in psycholinguistic tasks. However, in the case of constraints on Turkish vowels, it is a necessary condition that the constraint in question be derived from an alternation. Even alternations that have large numbers of lexical exceptions have have robust effects on a constrained wordlikeness task. Purely statistical approaches to phonotactic knowledge fail to draw this important distinguish.

>>>>

There are several arguments against this inference. First, there are many cases of statistical generalizations with a known diachronic cause but lacking any synchronic reality. One of the earliest arguments of this type was put forth by Saussure (\citeyear[202f.]{CLG}), who recognizes the a diachronic process which changed Old Latin intervocalic \emph{s} to \emph{r}, and that Classical Latin intervocalic \emph{s} is rare as a result. However, Saussure denies that this generalization has any status in Classical Latin.\footnote{See \citet{Gorman2012e} and citations therein for a full explication of Saussure's position on rhotacism, and a restatement of the phonological argument in generative terms.} Similarly, [ʃ] in Modern English derives from Old English [sk] (e.g., \emph{fisc} 'fish'), and since long vowels before complex codas are not found in Old English, [ʃ] is rarely preceded by long vowels in word-final syllables to this day. \citet{Iverson2005} label this constraint ``phonologically accidental'', since a millennium of speakers have produced coinages (e.g., \emph{posh}, \emph{swoosh}) and borrowings (e.g., \emph{douche}, \emph{capiche}) in utter disregard of this generalization. Secondly, wordlikeness generalizations are absolutely constrained by negative results in learnability (see \citealt{Yang2012} for a recent review). Finally, there are a number of results which suggest patterns which were obvious to the linguist go unlearned \citep[e.g.,][]{Becker2011,Hayes2006,Hayes2009,HayesInPress}. This argues against a direct link between statistical criteria and the synchronic reality of phonotactic generalizations. The danger is that grammatical models which closely mimic lexical statistics but which lack psycholinguistic support (e.g., Gitksan: \citealt{Brown2010}; Muna: \citealt{Anttila2008a}, \citealt{Coetzee2008a}; Shona: \citealt[][385]{Hayes2008a}) may overstate the phonotactic generalizations internalized by speakers.

This chapter is a comprehensive investigation of three well-known phonotactic generalizations in Turkish using lexical statistics and the results of a wordlikeness task performed by \citet{Zimmer1969}. \citeauthor{Zimmer1969}'s experiments merit reconsideration because there is widespread disagreement concerning the results and prior discussions do not relate behavioral data to lexical statistics and to competing formalizations of the generalizations involved.

@@

Two less familiar notations are used in this chapter. First, directional application conditions are used to derive the left-to-right, iterative properties of the harmony rules. Directional application explains provides a panacea for the considerable body of evidence against simultaneous rule application adduced in the 1970s (e.g., \citealt[209f.]{Anderson1974}, \citealt{Howard1972}, \citeauthor{Kenstowicz1973} \citeyear[14f.]{Kenstowicz1973}, \citeyear[189f.]{Kenstowicz1977}, \citeyear[318f.]{Kenstowicz1979}, \citealt{Piggott1975}). In fact, harmony in Turkish is one such case, insofar as simultaneous application makes the wrong predictions. Further, \citet{Johnson1972} and \citet{Kaplan1994} prove that rules specified for directional application are no more complex in the terms of formal learnability than that do not. Secondly, rather than the use of an unbounded number of Greek-letter variables ($\alpha$, $\beta$, etc.) over feature values $\{+, -\}$, only a single variable, denoted by `$=$', is used \citep{McCawley1973}. A structural description [$=$F]\ldots{}[$=$F] matches a string S$_i$\ldots{}S$_j$ if and only if S$_i$ and S$_j$ are both [$+$F] or both [$-$F].\footnote{This is more restrictive than the Greek-letter variable notation, in that the value of $=$ for some feature F cannot be applied to F$'$ if F $\ne$ F$'$; \citet{Odden2012} dismisses well-known counterexamples against this restriction adduced in \emph{SPE} (352-353).}

@@

It is important to dispense with an alternative proposed by \citet{Clements1982} and \citet{Inkelas1997} which reintroduces this duplication. Root vowels exhibit a robust contrast for backness (e.g., \emph{kül} `ash' vs.  \emph{kul} `servant', \emph{kepek} `bran' vs. \emph{kapak} `lid'), whereas harmonic roots are those in which the backness of any remaining vowels is predictable. \citeauthor{Clements1982} propose that these vowels, as well as harmonizing suffix vowels, are underspecified for backness, whereas the non-initial vowels of disharmonic roots are fully specified. This is schematized below.

@@

While harmony in non-derived environments can be inferred from the aforementioned suffix alternations, no evidence has yet been presented to show that Turkish speakers internalize the tendency for roots to conform to backness harmony. If Turkish speakers do not attend to this generalization, there is no need for the grammar to account for it. Several other ``external'' facts suggest  that this is not the case, with the caveat that uncritical acceptance of hypotheses linking loanword adaptation and language games to the grammar is no way being implied.

@@

A number of studies have investigated the role of harmony in word-spotting tasks, though to mimic auditory word recognition and segmentation in natural settings. Many of these studies have been carried out in Finnish, which has a vowel harmony system similar to that of Turkish. \citet{Suomi1997} and \citet{Vroomen1998} task Finnish speakers with identifying harmonic disyllables in an auditory stream. When the syllable preceding the disyllabic target has a different backness specification than the target, recognition of the target is facilitated. Presumably, disharmony facilitates the recognition of word boundaries. \citet{Kabak2010} find that Turkish \textsc{Backness Harmony} has a a similar effect: Turkish speakers are quicker and more accurate at the task of spotting the nonce target word \emph{pavo} when preceded by a disharmonic juncture (e.g., \emph{gölü-PAVO}) than when preceded by a harmonic juncture (e.g., \emph{golu-PAVO}). \citet{Kabak2010} find that this effect does not obtain for speakers of French, a language which lacks vowel harmony. This implies that Turkish speakers have internalized the tendency of harmonic sequences to be root-internal and of disharmonic transitions to cross word boundaries.

@@

\noindent 
This claim can be evaluated using the Fisher exact test. Let P denote a sequence of one or more consonants, one of which is labial, and let T denote a sequence of one or more consonants none of which is labial. The null hypothesis is that \emph{a}P\emph{u} sequences, with P representing a cluster containing a labial consonant, which conform to \textsc{Labial Attraction}, are no more likely than would be expected from other \emph{a}T\emph{u} sequences violating \textsc{Roundness Harmony}. The resulting counts are shown in Table \ref{las}. Whereas the sequence \emph{a}P\emph{u} is more than twice as likely as \emph{a}P\emph{ı}, the sequence \emph{a}T\emph{u} is four times less likely than \emph{a}T\emph{ı}. This is a significant interaction predicted by \textsc{Labial Harmony}, but contradicting a statistical interpretation of the claim by \citeauthor{Clements1982}.

@@

Both of the \citet{Zimmer1969} experiments include 5 pairs which differ on in whether or not the nonce words conform to, or violate, \textsc{Backness Harmony}. As can be seen from Table \ref{bhw}, harmonic pairs are preferred approximately 6-to-1, and aggregating over speakers, no disharmonic member of a pair is favored. Speakers have a highly reliable preference for nonce words which exhibit \textsc{Backness Harmony} ($\gamma = 0.694$, $p = 1.7$\e{-59}). It is interesting to note that the disharmonic nonce word which has the highest rating is found in the pair \emph{terüz}-\emph{teruz}, both of which violate \textsc{Roundness Harmony}. While this is little more than an anecdote, this may be  indicative of a link between the two processes, and their exceptions, in the minds of native speakers.

@@

\citet{NiChiosain1993} and \citet{Ito1995b} argue that \textsc{Labial Attraction} is true only of native vocabulary. This would be a potential confound for the experimental results, since it is not implausible that the subjects' in \citeauthor{Zimmer1969}'s study treated nonce words as if they were loanwords. 
%\footnote{Indeed, many wordlikeness studies include instructions to the participants to treat the stimuli much as if they were loanwords \citep[e.g.,][]{Hay2004a}. This may influence the choice of task model used by participants.}
This was not the opinion of \citet[266]{Lees1966a}, and it is conclusively refuted by  \citet{Inkelas2001}. They observe that in the subset of the TELL database with etymological coding, a full 75\% of foreign vocabulary conform to \textsc{Labial Attraction}. Among native lexical items, only 52\% do, a significant trend but in an unexpected direction (Fisher exact test, $p = 0.042$). One plausible explanation is that many of the languages in contact with Turkish lack the \emph{ı} phoneme needed to violate \textsc{Labial Attraction}.

@@

It has been shown that statistical reliability is not a sufficient condition for a sequence structure generalization to be reflected in psycholinguistic tasks. However, in the case of constraints on Turkish vowels, it is a necessary condition that the constraint in question be derived from an alternation. Even alternations that have large numbers of lexical exceptions have robust effects on a constrained wordlikeness task. Purely statistical approaches to phonotactic knowledge fail to draw this important distinction.
